# Crop domestication, origin, germplasm resource types

For many of the more widely grown food crops, domestication resulted in evolutionary changes making them genetically distinct from their closest wild relatives today, and most became dependent on humans for reproduction (Harlan, 1992; Simmonds and Smartt, 1999).

Domestication seems likely to have been the result of indirect selection and unintentional direct selection (e.g. when farmers select for large seed size or brittle rachis as a result of their seed collection behaviour; Harlan, 1992), and perhaps some intentional selection for evolutionary change. However, it is very difficult or impossible to determine the type of selection that resulted in past crop evolution. Allard argue of direct, intentional selection expressing the view that early farmers were competent biologists. In contrast, Simmonds and Smartt emphasize indirect selection.

Similar to studies based on archaeological data, results of molecular analyses support the hypothesis that farmers' selection has been successful in achieving evolutionary change for traits in the 'domestication syndrome' that might be indirectly or unintentionally favoured because of agronomic superiority as well as other preferences. For example, three major genes involved in starch metabolism in maize were found to have unusually low genetic diversity compared with its closest wild relative (teosinte, _Zea mays_ subsp. _parviglumis_), which is strong evidence of selection for specific processing and culinary qualities important for the primary manner in which maize has been consumed in its regions of origin and diversity.

Similarly, it appears that strong directional selection for sticky, glutinous grain quality resulted in a selective sweep affecting an area over 250 kb long that includes the locus coding for this quality (low amylase production) and other linked loci. The presence of this sweep distinguishes the sticky rice favoured by upland northeast Asian peoples from the non-glutinous rice varieties used by other Asian groups, and presumably would be among their fundamental choice criteria, perhaps as an adaptation for eating with chopsticks.

Increasing evidence for a number of crops suggests that domestication could have occurred over short periods relative to the ~12 000 years that crop plants have been cultivated (Gepts, 2004). Domestication syndrome traits often appear to be determined by a small number of genes with large effects, suggesting that domestication could proceed relatively rapidly. In common bean ( _Phaseolus vulgaris_ L.), control of the domestication syndrome involves genes that have a large effect (>25–30 percent) and account for a substantial part of the phenotypic variation observed (>40–50 percent) (Koinange, Singh and Gepts, 1996).

Domestication in sexually propagated crops may have resulted in increased autogamy and therefore homozygosity, expressed phenotypically in greater trueness to type in a population over generations. In contrast, some vegetative propagation may have selected for heterozygosity (via heterosis) and therefore for allogamy, as contemporary evidence suggests for cassava.

Domestication generally decreased the fitness of plants in natural environments, and made them more dependent on humans and human-managed environments. The geographical spread of domesticated crops led to great varietal diversification as a result of the increase in diversity of natural and artificial selection pressures encountered, followed by choice among preferred populations.

## Center of origin and estimated time of cultivation

```{r center-of-origin-cultivation}
tribble(
  ~"Crop",	~"Length of time domesticated (years)",	~"Possible region of origin",
  "Maize, Zea mays",	"7000",	"Mexico, Central America",
  "Rice, Oryza sativa",	"4500",	"Thailand, Southern China",
  "Wheat, Triticum spp.",	"8500",	"Syria, Jordan, Israel, Iraq",
  "Barley, Hordeum vulgare",	"9000",	"Syria, Jordan, Israel, Iraq",
  "Sorghum, Sorghum bicolor",	"8000",	"Equatorial Africa",
  "Soybean, Glycine max",	"2000",	"North China",
  "Oil palm, Elaeis guineensis",	"9000", "Central Africa",
  "Coconut palm, Cocos nucifera",	"100", "Southern Asia",
  "Rapeseed, Brassica napus",	"500",	"Mediterranean Europe",
  "Sunflower, Helianthus annus",	"3000",	"Western United States",
  "Beans, Phaseolus spp",	"7000",	"Centra America, Mexico",
  "Lentil, Lens culinaris",	"7000",	"Syria, Jordan, Israel, Iraq",
  "Peas, Pisum sativum",	"9000",	"Syria, Jordan, Israel, Iraq",
  "Potato, Solanum tuberosum",	"7000",	"Peru",
  "Cassava, Manihot esculenta",	"5000",	"Brazil, Mexico",
  "Sweet potato, Ipomoea batatas",	"6000",	"South Central America",
  "Sugar beet, Beta vulgaris",	"300",	"Mediterranean Europe",
  "Tomato, Lycopersicum esculentum",	"3000",	"Western South America",
  "Cabbage, Brassica oleracea",	"3000",	"Mediterranean Europe",
  "Onion, Allium spp.",	"4500",	"Iran, Afganistan, Pakistan",
  "Orange, Citrus sinensis",	"9000",	"South-east Asia",
  "Apple, Malus spp.",	"3000",	"Asia Minor, Central Asia",
  "Grape, Vitis spp.",	"7000",	"Eastern Asia",
  "Banana, Musa acuminata, M. balbisiana",	"4500",	"South-east Asia",
  "Cotton, Gossypium spp.",	"4500",	"Centra America, Brazil",
  "Coffee, Coffea spp.",	"500",	"West Ethiopia",
  "Rubber, Hevea brasiliensis",	"200",	"Brazil, Bolivia, Paraguay",
  "Alfalfa, Medicago sativa",	"4000",	"Iran, Northern Pakistan",
) %>% 
  knitr::kable(booktabs = TRUE, longtable = TRUE) %>% 
  kableExtra::kable_styling(position = "center", latex_options = "striped", font_size = 8) %>% 
  kableExtra::column_spec(column = 1:3, width = c("14em", "12em", "12em"))

```


### Crops of indian origin

```{r indian-main-center-origin}
indian_origin_crop <- bind_rows(
  read_csv("./data/agronomy_crops_origin_indian_main_center1.csv", col_names = c("Group", "Crops")) %>% 
    tidyr::fill(Group, .direction = "down"),
  read_csv("./data/agronomy_crops_origin_indian_main_center2.csv", col_names = c("Group", "Crops")) %>% 
    tidyr::fill(Group, .direction = "down"))

indian_origin_crop %>% 
  separate(col = "Crops", sep = ", ", into = c("Common name", "Scientific name")) %>% 
  mutate(`Scientific name` = paste0("\\textit{", `Scientific name`, "}")) %>% 
  knitr::kable(booktabs = TRUE, caption = "Crops with their origin in Indian Main center", escape = FALSE, longtable = TRUE) %>% 
  # kableExtra::kable_styling(font_size = 8, latex_options = "striped") %>%
  kableExtra::kable_styling(font_size = 8) %>% 
  kableExtra::column_spec(column = 1:3, width = c("8em", "8em", "15em")) %>% 
  kableExtra::collapse_rows(1:2, row_group_label_position = "stack")
```

```{r indian-indo-malayan-origin}
read_csv("./data/agronomy_crops_origin_indo_malayan_center.csv", col_names = c("Group", "Crops")) %>% 
  tidyr::fill(Group, .direction = "down") %>% 
  separate(col = "Crops", sep = ", ", into = c("Common name", "Scientific name")) %>% 
  mutate(`Scientific name` = paste0("\\textit{", `Scientific name`, "}")) %>% 
  knitr::kable(booktabs = TRUE, caption = "Crops with their origin in Indo-Malyan center", escape = FALSE, longtable = TRUE) %>% 
  # kableExtra::kable_styling(font_size = 8, latex_options = "striped") %>% 
  kableExtra::kable_styling(font_size = 8) %>% 
  kableExtra::column_spec(column = 1:3, width = c("8em", "8em", "15em")) %>% 
  kableExtra::collapse_rows(1:2, row_group_label_position = "stack")
```

```{r indian-central-asiatic-origin}
read_csv("./data/agronomy_crops_origin_central_asiatic_center.csv", col_names = c("Group", "Crops")) %>%
  tidyr::fill(Group, .direction = "down") %>%
  separate(col = "Crops", sep = ", ", into = c("Common name", "Scientific name")) %>%
  mutate(`Scientific name` = paste0("\\textit{", `Scientific name`, "}")) %>%
  knitr::kable(booktabs = TRUE, caption = "Crops with their origin in Central Asiatic center", escape = FALSE, longtable = TRUE) %>%
  # kableExtra::kable_styling(font_size = 8, latex_options = "striped") %>%
  kableExtra::kable_styling(font_size = 8) %>%
  kableExtra::column_spec(column = 1:3, width = c("6em", "8em", "15em")) %>%
  kableExtra::collapse_rows(1:2, row_group_label_position = "stack", longtable_clean_cut = FALSE)
```

## Germplasm resources

### Gene pool concept

A spectrum of different vegetables could be derived from the same wild progenitor. Brassica oleracea stands as an excellent example for this biological process. Wild progenitor is a weedy herb that grows on limestone in the Mediterranean region. Domestication of several distinct lineages of _B. oleracea_ produced several vegetable varieties or cultivar groups or subspecies ("ssp."): kale and collard greens (ssp. acephala), Chinese broccoli (ssp. alboglabra), red and green cabbages (ssp. capitata), savoy cabbage (ssp. sabauda), kohlrabi (ssp. gongylodes), Brussels sprouts (ssp. gemmifera), broccoli (ssp. italica) and cauliflower (ssp. botrytis). Though these varieties look dramatically different, they are considered the same species since they are all inter-fertile, capable of mating with one another and producing fertile offspring.

### Types
